---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)

description <- read.dcf("DESCRIPTION", fields = c("Version", "Depends"))
package_version <- description[1, "Version"]
r_dependencies <- trimws(strsplit(description[1, "Depends"], ",")[[1]])
r_requirement <- r_dependencies[startsWith(r_dependencies, "R ")]
minimum_r_version <- sub("^R \\(>= ([^)]+)\\)$", "\\1", r_requirement)
stopifnot(length(minimum_r_version) == 1L)
```

# rDGIdb

<!-- badges: start -->
[![Code QA](https://github.com/dgidb/r-dgidb/actions/workflows/code_qa.yaml/badge.svg)](https://github.com/dgidb/r-dgidb/actions/workflows/code_qa.yaml)
[![R-universe testing](https://github.com/dgidb/r-dgidb/actions/workflows/r-universe.yml/badge.svg)](https://github.com/dgidb/r-dgidb/actions/workflows/r-universe.yml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](LICENSE)
<!-- badges: end -->

`rDGIdb` provides an R interface to the
[Drug-Gene Interaction Database (DGIdb)](https://dgidb.org/) GraphQL API.
It supports queries for drugs, genes, drug-gene interactions and their claim
types, gene categories, DGIdb data sources, and Drugs\@FDA application data.

## Installation

The current development version is `r package_version` and requires
R `r minimum_r_version` or later.

The development version can be installed from GitHub with:

```{r install-dev, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("dgidb/r-dgidb")
```

Alternatively, after cloning the repository, install the local checkout with:

```{r install-local, eval = FALSE}
devtools::install(build_vignettes = TRUE)
```

## Quick start

Load the package:

```{r load}
library(rDGIdb)
```

Query functions return base R data frames. Fields that can contain multiple
values for one record, such as aliases or interaction sources, are represented
as list-columns.

Look up genes:

```{r genes, eval = FALSE}
genes <- getGenes(c("BRAF", "PDGFRA"))
genes
```

Retrieve drug-gene interactions for genes of interest:

```{r interactions, eval = FALSE}
interactions <- getInteractions(c("BRAF", "PDGFRA"))
interactions
```

Look up a drug:

```{r drugs, eval = FALSE}
drugs <- getDrugs("Imatinib")
drugs
```

Interactions can also be queried starting from a drug:

```{r drug-interactions, eval = FALSE}
getInteractions(
    "Imatinib",
    search = "drugs",
    approved = TRUE
)
```

Retrieve the current interaction-type vocabulary and use a returned value as
an interaction filter:

```{r interaction-types, eval = FALSE}
interaction_types <- getInteractionTypes()
getInteractions(
    "BRAF",
    interactionType = interaction_types$interaction_type[[1]]
)
```

Other package functions provide access to DGIdb gene categories, source
databases, complete gene and drug lists, and Drugs\@FDA application data.

```{r other-functions, eval = FALSE}
getCategories("BRAF")
getSources(sourceTypes$INTERACTION)
getInteractionTypes()
getDrugApplications("Imatinib")
getAllGenes()
getAllDrugs()
```

### Enriching drugs with Drugs\@FDA data

`getDrugApplications()` combines application identifiers from DGIdb with
product-level records from the openFDA Drugs\@FDA API. It returns one row per
product, keeps active ingredient names and strengths together in the
`fda_active_ingredients` list-column, and records unsuccessful lookups in
`fda_lookup_status` and `fda_lookup_message` instead of silently dropping them.

```{r fda-applications, eval = FALSE}
applications <- getDrugApplications("Imatinib")
applications[c(
    "drug_name", "fda_application_number", "fda_brand_name",
    "fda_marketing_status", "fda_lookup_status"
)]

applications$fda_active_ingredients[[1]]
```

An openFDA API key is optional. For regular or larger workloads, set it once in
the `OPENFDA_API_KEY` environment variable, or pass it with the `fdaApiKey`
argument. Avoid placing API keys directly in scripts committed to version
control.

```{r fda-key, eval = FALSE}
Sys.setenv(OPENFDA_API_KEY = "your-api-key")
getDrugApplications(c("Imatinib", "Gefitinib"))
```

The function batches repeated application lookups within each call. Drugs\@FDA
data are intended for research and should not be the sole basis for medical
decisions.

For a complete walkthrough, see the
[introductory vignette](vignettes/rDGIdb.Rmd). With vignettes installed, open
it in R with:

```{r vignette, eval = FALSE}
vignette("rDGIdb")
```

## Migrating from the legacy interface

For replacements for `queryDGIdb()` and the legacy result accessors, see the
[migration guide](vignettes/migrating-from-legacy-rDGIdb.Rmd). With vignettes
installed, open it in R with:

```{r migration-vignette, eval = FALSE}
vignette("migrating-from-legacy-rDGIdb")
```

## Getting help

If you encounter a problem with `rDGIdb`, first check the package documentation
and vignette.

```{r help, eval = FALSE}
help(package = "rDGIdb")
vignette("rDGIdb")
```

Bug reports and feature requests can be submitted through the
[GitHub issue tracker](https://github.com/dgidb/r-dgidb/issues).

## Citation

To obtain the preferred citation for `rDGIdb`, run:

```{r citation, eval = FALSE}
citation("rDGIdb")
```

Please also cite the Drug-Gene Interaction Database (DGIdb) when appropriate.

## Code of Conduct

Please note that the `rDGIdb` project follows the
[Bioconductor Code of Conduct](https://bioconductor.org/about/code-of-conduct/).
By contributing to this project, you agree to abide by its terms.

## Development

See [CONTRIBUTING.md](CONTRIBUTING.md).
