---
title: "Migrating from legacy rDGIdb"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('rDGIdb')`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Migrating from legacy rDGIdb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
online <- interactive()
```

# Start with a query

This guide is for users of the legacy `rDGIdb` interface built around
`queryDGIdb()` and its result accessors. The new interface queries DGIdb through
GraphQL and returns base R data frames. Legacy functions and result classes
are not provided by the new interface.

Previously:

```{r legacy-query, eval = FALSE}
library(rDGIdb)
genes <- c("TNF", "AP1", "AP2", "XYZA")
result <- queryDGIdb(genes)
detailedResults(result)
```

Now:

```{r load-package}
library(rDGIdb)
```

```{r new-query, eval = online}
genes <- c("TNF", "AP1", "AP2", "XYZA")
interactions <- getInteractions(terms = genes)
head(interactions[c("gene_name", "drug_name", "interaction_score")])
```

The returned data frame is the result: there is no accessor to call first.
Its columns differ from the legacy detailed results. Multi-valued fields,
including `interaction_sources`, `interaction_pmids`, and
`interaction_attributes`, are list-columns. For example,
`interactions$interaction_sources[[1]]` contains the sources for the first
row, when a row is present. Interaction type labels are not a column in this
result, even though they can be used as query filters.

# Function reference

| Legacy function | New interface or recipe |
| --- | --- |
| `queryDGIdb(genes)` | `getInteractions(terms = genes)` |
| `detailedResults(result)` | Use the returned interaction data frame directly |
| `resultSummary(result)` | Count rows, unique gene IDs, and unique drug IDs; see below |
| `byGene(result)` | Split the interaction data frame by gene |
| `searchTermSummary(result)` | No direct replacement for legacy search-term match classifications |
| `sourceDatabases()` | `getSources(sourceTypes$INTERACTION)`; use `source_short_name` for filters |
| `geneCategories()` | `getCategories(genes)` returns annotations for specified genes, not a global vocabulary |
| `interactionTypes()` | `getInteractionTypes()`; extract unique, nonmissing `interaction_type` values |
| `plotInteractionsBySource(result)` | Count source membership and use base R `barplot()` |
| `resourceVersions()` | Inspect names and versions from `getSources()` |

The recipes below describe their counting and filtering rules explicitly.
They do not reproduce the legacy result object's exact structure or guarantee
identical historical results.

# Filters and available values

## Source databases and interaction types

The basic argument changes are:

| Legacy argument | New argument |
| --- | --- |
| `genes` | `terms` (with the default `search = "genes"`) |
| `sourceDatabases` | `source` |
| `interactionTypes` | `interactionType` |
| `geneCategories` | Retrieve annotations separately and subset locally, as below |

Omit optional filters or use `NULL` to leave them unset. The current GraphQL
queries accept a **single string** for `source` and `interactionType`.
Legacy examples passing vectors of sources or types are therefore not direct
replacements. Choose one value per filter for a query; combining separate
queries would require defining union/intersection and deduplication rules for
your analysis.

```{r filter-values, eval = online}
sources <- getSources(sourceTypes$INTERACTION)
sources[c("source_name", "source_short_name")]

interaction_types <- getInteractionTypes()
type_values <- unique(na.omit(interaction_types$interaction_type))
type_values
```

Use `source_short_name`, which contains the database identifiers used by
queries, rather than the display names in `source_name`. Multiple interaction
type records can have the same label; `unique()` above gives filter values
without discarding records from the original metadata table.

For example, using one of each available value:

```{r filtered-query, eval = online}
if (nrow(sources) > 0L && length(type_values) > 0L) {
    filtered <- getInteractions(
        "BRAF",
        source = sources$source_short_name[[1]],
        interactionType = type_values[[1]]
    )
}
```

That combination may return no interactions. Available values and underlying
DGIdb data can change over time.

## Gene categories

`getCategories()` requires gene names and returns gene-category annotations.
It does not replace the legacy zero-argument vocabulary lookup. Inspect
`unique(categories$gene_category)` to find categories among the queried genes.
To filter interactions by category, match the annotated gene concept IDs:

```{r category-query, eval = online}
categories <- getCategories(genes)
selected_categories <- "CLINICALLY ACTIONABLE"
selected_ids <- unique(categories$gene_concept_id[
    categories$gene_category %in% selected_categories
])
category_interactions <- interactions[
    interactions$gene_concept_id %in% selected_ids, ,
    drop = FALSE
]
```

This retains interactions for genes with any selected category. A gene without
a matching annotation is excluded; an empty selection produces a zero-row
data frame. Check the returned category labels before choosing values.

# Working with results

The following small data sets are **illustrative**, not retrieved DGIdb
records. They let the recipes run without network access and include repeated
source claims to show how counting works. Apply the same operations to your
own `interactions` and `categories` data frames.

```{r illustrative-data}
example_interactions <- data.frame(
    gene_name = c("GENE_A", "GENE_A", "GENE_B"),
    gene_concept_id = c("gene:a", "gene:a", "gene:b"),
    drug_name = c("DRUG_X", "DRUG_Y", "DRUG_X"),
    drug_concept_id = c("drug:x", "drug:y", "drug:x")
)
example_interactions$interaction_sources <- list(
    c("Source A", "Source A", "Source B"),
    "Source A",
    character()
)
example_categories <- data.frame(
    gene_concept_id = c("gene:a", "gene:b"),
    gene_category = c("CATEGORY_1", "CATEGORY_2")
)

selected_ids <- unique(example_categories$gene_concept_id[
    example_categories$gene_category %in% "CATEGORY_1"
])
example_selected <- example_interactions[
    example_interactions$gene_concept_id %in% selected_ids, ,
    drop = FALSE
]
example_selected
```

## Summaries and grouping

For a replacement for common `resultSummary()` tasks, state what is counted:

```{r summarize-results}
summarize_interactions <- function(x) {
    c(
        interaction_rows = nrow(x),
        genes = length(unique(x$gene_concept_id)),
        drugs = length(unique(x$drug_concept_id))
    )
}
summarize_interactions(example_interactions)
```

These are counts of returned rows and distinct concept identifiers, not counts
of input search terms or evidence claims. A simple alternative to `byGene()`
is a named list of data frames:

```{r group-results}
by_gene <- split(example_interactions, example_interactions$gene_name)
by_gene[["GENE_A"]]
```

## Plot interactions by source

Each source counts at most once per interaction row. An interaction supported
by two sources contributes to both bars, so the sum can exceed the number of
rows. Rows with no sources contribute to no bars.

```{r plot-sources, fig.width = 7, fig.height = 4}
count_sources <- function(x) {
    source_values <- unlist(
        lapply(x$interaction_sources, unique),
        use.names = FALSE
    )
    sort(table(source_values), decreasing = TRUE)
}
source_counts <- count_sources(example_interactions)
if (length(source_counts) > 0L) {
    barplot(
        source_counts,
        main = "Number of interaction rows by source",
        ylab = "Interaction rows"
    )
}
```

## Empty results and search terms

Queries with no returned interactions produce a zero-row data frame with the
usual columns. The summaries above return zeros, grouping returns an empty
list, and the plot guard skips an empty chart:

```{r empty-results}
empty_interactions <- example_interactions[FALSE, , drop = FALSE]
summarize_interactions(empty_interactions)
split(empty_interactions, empty_interactions$gene_name)
count_sources(empty_interactions)
```

There is no direct replacement for `searchTermSummary()`. A gene absent from
interaction results may have no interactions under the selected filters;
absence does not prove that its input name was unrecognized. `getGenes()` can
be used for a separate record lookup, but its data frame does not expose the
legacy per-input ambiguous/unmatched classifications. Comparing input strings
to returned gene names also cannot reconstruct those classifications, because
returned names may differ from the supplied terms.

# Source versions and help

For information previously obtained through `resourceVersions()`, inspect
source metadata:

```{r source-versions, eval = online}
source_versions <- getSources()
source_versions[c("source_name", "source_short_name", "source_version")]
```

Record these values alongside your query and analysis date when useful. They
are metadata from the current service, not a way to select an older snapshot.

Use the new function help pages in place of `?queryDGIdb`, `?rDGIdbFilters`,
`?rDGIdbResult`, and `?plotInteractionsBySource`:

```{r help, eval = FALSE}
?getInteractions
?getCategories
?getSources
?getInteractionTypes
vignette("rDGIdb")
citation("rDGIdb")
```

The introductory vignette covers additional capabilities, including drug
lookups and Drugs@FDA application information. Upstream steps that derive gene
names from variant annotations can still supply their character vector to
`getInteractions(terms = genes)`.

# Session information

```{r session-info}
sessionInfo()
```
